---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(ctypesio)
```

# ctypesio

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![CRAN](http://www.r-pkg.org/badges/version/ctypesio)](https://cran.r-project.org/package=ctypesio)
[![R-CMD-check](https://github.com/coolbutuseless/ctypesio/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/ctypesio/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{ctypesio}` is a package fore reading and writing standard C types from connections or raw vectors.

Motivation:  I wanted to be able to read/write some binary data files not currently 
supported by R.  I needed more robust error-handling, compatibility with 
C types and a better experience than just using `readBin()` directly.

Key features:

* Read 8, 16, 32, 64 signed and unsigned integers
* Read 16, 32 and 64bit floating point values
* Support for integer types which contain values outside the range of
  R's integer type
* Persistent information is stored on the connection for how to handle:
    * endianness
    * bounds checking
    * EOF behaviour
    * `NA` checking

## What's in the box

* Floating point numbers
    * read/write double-, single- and half-precision floats
    * `read_f64()`, `write_f64()` `read_f32()`, `write_f32()`, `read_f16()`, `write_f16()`
    * Aliased functions for: `read_dbl()`, `write_dbl()`, `read_float()`, `write_float()`, 
      `read_half()`, `write_half()`
    * `read_bfloat()` read 16-bit [bfloat](https://en.wikipedia.org/wiki/Bfloat16_floating-point_format) numbers
* Integers
    * read/write signed/unsigned 8, 16, 32, 64-bit integers
    * `read_uint8()`, `write_uint8()`, `read_uint16()`, `write_uint16()`, `read_uint16()`, `write_uint16()`,
    `read_uint16()`, `write_uint16()`
      integers
    * For types which don't fit in a standard R integer (e.g. `uint64`) values can be promoted to:
        * double
        * hexadecimal string
        * `bit64::integer64` (this is a signed 64bit integer)
        * raw vector
* String handling
    * `read_str()`, `read_str_raw()`, `read_utf8()`, `read_utf8_raw()`, 
       `write_utf8()`, `write_utf8_raw()`
    * `scan_dbl()`, `scan_int()`, `scan_str()` are wrappers around 
      R's `scan()` 
    * `fprintf()`, `fprintf_raw()`
    * Functions with the suffix `_raw` read/write strings which are not
      nul-terminated
* Raw bytes       
    * `read_raw()`, `write_raw()` Read/write raw bytes
    * `read_hex()`, `write_hex()` Read/write bytes as hex strings
* Data layout operations
    * `flip_endian()` reverse the elements within blocks within a vector. Used
      for changing *endianness* of data in a vector.
    * `aperm_array_to_vector()` and `aperm_vector_to_array()` are wrappers around
      `aperm()` to allow for easier import/export of array data.
* Connection configuration
    * Set persistent configuration operations on the connection so that they 
      do not need to be re-specified for every read/write operation i.e.
        * Endianness
        * Bounds checking
        * EOF behaviour
        * NA checking
    * `set_endian()` set the default *endianness* for all operations on this
       connection.  Can be over-ridden within each function by specifying 
       `endian` argument. If not specified on the connectino or in 
       function argument, this defaults to `little` on
       all platforms.
    * `set_bounds_check()` set handling of out-of-bounds checks when 
       writing integer/float values.
        * `ignore` do nothing and let underlying `writeBin()` truncate data
          however it wants
        * `warn` warn when data is outside the bounds for this type. E.g.
          trying to write an integer value of `500` with `write_uint8()`
          will trigger this warning.
        * `error` same as `warn` except an actual error is raised.
    * `set_eof_check()` check if the number of items returned does not match
      the number requested. This would indicate the connection reached EOF
      during data reading.  Options are to `ignore`, `warn` or `error`
    * `set_na_check()` check for NAs in data prior to writing.
      Options are to `ignore`, `warn` or `error`

| bits | integer                                                                  | floating point                      |
|------|--------------------------------------------------------------------------|-------------------------------------|
| 8    | Read/write. Signed/unsigned.                                             | NA                                  |
| 16   | Read/write. Signed/unsigned.                                             | Yes. Half-precision IEEE and bfloat |
| 32   | Read/write. Signed/unsigned. (32bit unsigned integer promoted to double) | Yes                                 |
| 64   | Read/write. Signed/unsigned. Double, hex, `bit64::integer64`, raw vectors | Yes                                 |

## Installation


This package can be installed from CRAN

``` r
install.packages('ctypesio')
```

You can install the latest development version from
[GitHub](https://github.com/coolbutuseless/ctypesio) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/ctypesio')
```

Pre-built source/binary versions can also be installed from
[R-universe](https://r-universe.dev)

``` r
install.packages('ctypesio', repos = c('https://coolbutuseless.r-universe.dev', 'https://cloud.r-project.org'))
```

## Example: Parsing a JPEG file

JPEG files are a sequence of chunks (identified by **marker** bytes), followed 
by a length and then a sequence of bytes.

These chunks continue until the `Start of Scan` chunk which contains the 
compressed data continuing until the length of the file.


```{r jpeg}
library(ctypesio)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Read the old logo from the 'jpeg' package
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
jpeg_file <- system.file("img", "Rlogo.jpg", package="jpeg")
jpeg <- jpeg::readJPEG(jpeg_file)
plot(as.raster(jpeg))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# What does the raw data look like?
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
dat <- readBin(jpeg_file, raw(), n = file.size(jpeg_file)) 
head(dat, 100)
```

#### Parsing the JPEG file

* Open the connection to the file
* Set the default endianness for all operations to *big endian*
* Confirm the file contains the SOI marker 'ffd8'
* Read the first marker and decode some information about the image.


JPEG files mostly contain *big-endian* encoded markers and integers, so 
we'll set this as the default on the connection (so we don't have to 
specify it for every read operation).

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Open a connection, and tag the connection such that 
# values are read in **big endian** by default.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
jpeg_file <- system.file("img", "Rlogo.jpg", package="jpeg")
con <- file(jpeg_file, 'rb') |>
  set_endian('big')

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Read the first 2 bytes as HEX
# For JPEG files, this should be the "Start of Image (SOI)" marker "ffd8"
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
soi <- read_hex(con, n = 1, size = 2) 
stopifnot(soi == 'ffd8')

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Next marker should be start of JFIF.  Marker hex = 'ffe0'
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
marker <- read_hex(con, n = 1, size = 2)
stopifnot(marker == 'ffe0')

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Data expected in this chunk
#  - length of this chunk    (16 bit unsigned int)
#  - the word 'JFIF'         (nul-terminated string)
#  - Major version number    ( 8 bit unsigned int)
#  - Minor version number    ( 8 bit unsigned int)
#  - Density units:          ( 8 bit unsigned int)
#       0 - no unit
#       1 - pixels-per-inch
#       2 - pixels-per-cm
#  - Xdensity                (16 bit unsigned int)
#  - Ydensity                (16 bit unsigned int)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
list(
  length   = read_uint16(con),
  jfif     = read_str(con),
  version  = read_uint8(con, 2),
  units    = c('', 'PPI', 'PPcm')[read_uint8(con) + 1],
  density = read_uint16(con, 2)
)
```

This information reveals that the JPEG is:

* a JFIF image
* Version 1.1
* with pixel density 300x300 dpi


```{r echo=FALSE}
close(con)
```



