---
title: "Handcrafting a WAV file"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handcrafting a WAV file}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ctypesio)
```

## WAV files

WAV files are a standard format for storing audio data.

This vignette will use the WAV file structure to write bespoke data as a 
WAV file, and then read it back in.


## WAV file structure

The 44-byte header structure for a simple wave file is shown below.

```{r echo=FALSE}
wav_info <- read.csv(textConnection(
r'{"Positions","Sample.Value","Description"
"1 - 4","'RIFF'","Marks the file as a riff file. Characters are each 1 byte long."
"5 - 8","File size (integer)","Size of the overall file - 8 bytes, in bytes (32-bit integer). Typically, you’d fill this in after creation."
"9 -12","“WAVE”","File Type Header. For our purposes, it always equals “WAVE”."
"13-16","'fmt '","Format chunk marker. Includes trailing space"
"17-20","16","Length of format data as listed above"
"21-22","1","Type of format (1 is PCM) - 2 byte integer"
"23-24","2","Number of Channels - 2 byte integer"
"25-28","44100","Sample Rate - 32 byte integer. Common values are 44100 (CD), 48000 (DAT). Sample Rate = Number of Samples per second, or Hertz."
"29-32","176400","(Sample Rate * BitsPerSample * Channels) / 8."
"33-34","4","(BitsPerSample * Channels) / 8.  (1)  8 bit mono (2) - 8 bit stereo/16 bit mono (4) - 16 bit stereo"
"35-36","16","Bits per sample"
"37-40","'data'","data chunk header. Marks the beginning of the data section."
"41-44","File size (data)","Size of the data section."
}'))

knitr::kable(wav_info, caption = "WAV file structure")
```

## Audio waveform

Create 2 seconds of random noise data to be played at 44.1 kHz.

Data values should be in range for a 16-bit integer i.e. [-32768, 32767]

```{r}
audio <- as.integer(runif(44100 * 2, -32768, 32767))
```

## Write the WAV file

Data is stored little-endian (which is the default endian mode for `ctypesio`)

```{r eval=TRUE}
wav_file <- tempfile(fileext = ".wav")
con <- file(wav_file, "wb")

write_utf8_raw(con, "RIFF")
write_uint32(con, length(audio) * 2 + 44 - 8) # total file length - 8
write_utf8_raw(con, "WAVE")
write_utf8_raw(con, "fmt ")
write_uint32(con, 16)                 # size of header so far
write_uint16(con, 1)                  # sample format = 2-byte integer
write_uint16(con, 1)                  # single channel
write_uint32(con, 44100)              # sample rate
write_uint32(con, 44100 * 16 * 1 / 8) # rate * bits * channels / 8
write_uint16(con, 2)                  # 16-bit mono audio
write_uint16(con, 16)                 # bits per sample
write_utf8_raw(con, 'data')
write_uint32(con, length(audio) * 2)  # sampled data size
write_int16(con, audio)

close(con)
```


## Play the audio

This audio isn't very interesting, but it can be played using the `audio` package

```{r eval=FALSE}
library(audio)
wav <- audio::load.wave(wav_file)
audio::play(wav)
```


# Parsing a WAV file

This is just the reverse of writing the file.

```{r}
con <- file(wav_file, "rb")

read_utf8_raw(con, 4)
read_uint32(con)      # total file length - 8
read_utf8_raw(con, 4)
read_utf8_raw(con, 4)
read_uint32(con)      # size of header above
read_uint16(con)      # sample format = 2-byte integer
read_uint16(con)      # single channel
read_uint32(con)      # sample rate
read_uint32(con)      # rate * bits * channels / 8
read_uint16(con)      # 16-bit mono audio
read_uint16(con)      # bits per sample
read_utf8_raw(con, 4)
(data_size <- read_uint32(con))         # sampled data size
data <- read_uint16(con, data_size / 2) # read in 16-bit samples

close(con)
```


## `ctypesio` is pipe-friendly

All the `write_*()` functions in `ctypesio` return the connection object, 
so that the functions can piped together.

```{r eval=FALSE}
wav_file <- tempfile(fileext = ".wav")
con <- file(wav_file, "wb")

write_utf8_raw(con, "RIFF") |>
  write_uint32(length(audio) * 2 + 44 - 8) |> # total file length - 8
  write_utf8_raw("WAVE")           |>
  write_utf8_raw("fmt ")           |> 
  write_uint32(16)                 |> # size of header so far
  write_uint16(1)                  |> # sample format = 2-byte integer
  write_uint16(1)                  |> # single channel
  write_uint32(44100)              |> # sample rate
  write_uint32(44100 * 16 * 1 / 8) |> # rate * bits * channels / 8
  write_uint16(2)                  |> # 16-bit mono audio
  write_uint16(16)                 |> # bits per sample
  write_utf8_raw('data')           |> 
  write_uint32(length(audio) * 2)  |> # sampled data size
  write_int16(audio)

close(con)
```



